\Huge{Complex State Management with Immutability}

### 2022-03-13 paper.tex
Linguistic advice taken into consideration

\title{\Huge{Complex State Management with Immutability}}
\author{Nassouh AlOlabi} \affil{Supervised By: Dr.Yasser Rahal, Fahmi Alammareen} 
\setlength{\affilsep}{1pt}
\date{}

@@

\vspace{1.5cm}

\noindent \textbf{Abstract}
\begin{center}
 	\line(1,0){430}
\end{center}
\vspace{-0,3cm}
\noindent Growing demand for fault-tolerant, scalable, distributed systems has made some mainstream software architectures and patterns obsolete or rather harder to come by, Thus came the rise of stateless and functional solutions based on data immutability which has already been the cornerstone of Big Data [1].

@@

\vspace{4mm}
\begin{center}
	\line(1,0){430}
\end{center}

@@

\subsection{Information Flow}
%%https://thedaylightstudio.com/blog/2018/03/14/what-is-state-in-web-application-development
Application State (a.k.a. Program State) represents everything necessary to keep your application running.  When we refer to application state we are normally referring to the state of the program as exists in the contents of its memory.  What does that mean practically?  How am I to understand that?  It helps to think in extremes.  What happens to information and functionality core to your application if a server goes down and restarts?  You lose whatever was residing in memory.

@@

% http://www.padsweb.rwth-aachen.de/wvdaalst/publications/p514.pdf
\subsubsection{Missing data}
Its' the situation where some data element needs to be accessed, i.e. read or destroyed, but either it has never been created or it has been deleted without having been created again.
\subsubsection{Inconsistent Data} 
Data is inconsistent if a task is using this data while some other task (or another instance of the same task) is writing to this data or is destroying it in parallel.
%-----------------------------
\subsection{Programming Paradigms}
% https://www.geeksforgeeks.org/introduction-of-programming-paradigms/
Programming paradigm is an approach to solve problem using some programming language or also we can say it is a method to solve a problem using tools and techniques that are available to us following some approach.
%----------------------------------------
And so most paradigms have their own opinion on data manipulation for example: Procedural \& Object Oriented paradigms allows more access to data modifying (destructive updates), deleting. Others like Logic \& Functional doesn't allow them.

\subsection{I/O, Network and Disk}
% https://www.solarwinds.com/-/media/solarwinds/swdcv2/licensed-products/storage-resource-monitor/resources/whitepapers/top_4_causes_of_storage_io_bottlenecks.ashx
Applications that are I/O heavy often cause bottlenecks. As well as often causing the problem, I/O intensive applications are often more sensitive to a storage latency issue. When you have a large user base trying to access these applications, slowdowns tend to take place. Increased response time in storage I/O causes bottlenecks. When there is a queue in the storage I/O, you would generally see an increase in latency.
Network and Disk latency are often times the biggest source of headache if poorly managed, which is often times the case since they're usually treated as if they were function calls which they're clearly not [3].
% http://norvig.com/21-days.html#answers
\begin{center}
    \begin{table}[H]
        \begin{center}
        \begin{tabularx} {0.8\textwidth}{ 
            | >{\raggedright\arraybackslash}X 
        | >{\raggedleft\arraybackslash}X | }
        \hline
        execute typical instruction &	1/1,000,000,000 sec = 1 nanosec\\
\hline
    fetch from L1 cache memory &	0.5 nanosec\\
\hline
branch misprediction &	5 nanosec\\
\hline
    fetch from L2 cache memory &	7 nanosec\\
\hline
    Mutex lock/unlock &	25 nanosec\\
\hline
    fetch from main memory &	100 nanosec\\
\hline
    send 2K bytes over 1Gbps network &	20,000 nanosec\\
    \hline
    read 1MB sequentially from memory &	250,000 nanosec\\
\hline
    fetch from new disk location (seek) &	8,000,000 nanosec\\
\hline
    read 1MB sequentially from disk &	20,000,000 nanosec\\
\hline
    send packet US to Europe and back &	150 milliseconds = 150,000,000 nanosec\\
    \hline      
    % \float
\end{tabularx}

\end{center}
\caption{ \label{tab:1} Latency numbers every programmer should know[4]}
\end{table}
\end{center}
As you can see disk fetches can take 100x more than main memory fetches. And when dealing with distributed systems were network call are an integral part of the system treating network call which 100000x more than main memory calls you end up with catastrophically unpredictable behavior

@@

% src
% Immutability an Empirical Study in Scala.pdf
When something is immutable, we say that it cannot be changed or that it is unchangeable. The definition of an immutable object in an object-oriented programming language is an object that has a state that cannot be mutated once instantiated (created). Moreover, an immutable class is a class whose instances cannot be mutated, meaning that there are no methods in the class that can mutate an instance of it. Classes that are not immutable can, however, have instances of it that are immutable.

\subsection{Forms of Immutability}
There are different varieties of immutability used in practice today, which includes:
\begin{itemize}
    \item Object immutability
    \begin{description}
        \item[] An immutable object is an object that cannot be modified (mutated), i.e., its state cannot be mutated
    \end{description}
    \item Class immutability 
    \begin{description}
        \item[] When every instantiated object of a class is immutable, then we say that the class itself is immutable.
    \end{description}
    \item Deep and shallow immutability (transitivity)\begin{description}
        \item[] Immutability can be deep or shallow, i.e., transitive or non-transitive:
        \item[]Deep (transitive): immutability means that all objects referred to by an immutable object must also be immutable.
        \item[]Shallow (non-transitive) immutability has a more relaxed constraint, and the immutable object may refer to objects that are mutable, but the fields of the immutable object itself cannot be mutated.
    \end{description}    
    \item Reference immutability (read-only references)
     \begin{itemize}
    \item[] Languages with the support of reference immutability have the notion of read-only references. A readonly reference cannot mutate the object it is referring to. When all references to an object are readonly, then nothing can mutate the object, and the object is immutable. There are, however, often no guarantee that a referred to object is immutable because the object may still be mutated by some other reference that is not read-only, unless there is some analysis that ensures that there only exist readonly references to that object. Reference immutability thus often ensure shallow (non-transitive) immutability and “reference immutability” does not mean that a reference itself is immutable [5].
\end{itemize}

\item Non-assignability
     \begin{itemize}
    \item[] Non-assignability is a form of immutability and property of a variable that makes sure that it cannot be reassigned. Since fields of objects are variables and if no variable can be reassigned after its initial assignment, the object is effectively immutable if what the fields are referring to is immutable. This make non-assignability give shallow (non-transitive) immutability too. Assignment of an object’s field mutates the object, but it does not mutate what was previously on the field or what was assigned to the field.
\end{itemize}

\item Concrete and abstract immutability
     \begin{itemize}
    \item[] Concrete immutability does not allow any change to the object’s state. Abstract immutability, however, allows an immutable object to change its internal state but not the object’s “abstract” value. The object is still immutable from the perspective of the object’s observer that can only see the abstract value, but the object may mutate internally. This can, for example, be useful to speed up certain operations, lazy initialization and buffering.
\end{itemize}

@@

\subsection{Properties of Immutability}
% write some problems
The concept of immutability is important and
utilizing immutable data is considered to ease software development and reasoning about programs in
numerous ways, for example:

\subsubsection{Predictability} It is harder to understand and reason about programs that have shared mutable
states with unclear interactions. Tracking mutations and maintaining the correct state of a program
can be difficult. Using immutable data naturally, avoids state changes and forces the programmer
to let data flow and be utilized in a different way throughout the program, making the state of the
program more predictable. For example, calling the same function twice would yield the same result,
and the outcome is predictable. Without the immutability guarantee, the second call could
yield another result because of an underlying state mutation making it less predictable.

\subsubsection{Testability} Because immutable data can only be changed once during construction, they are inherently
simpler and easier to unit test. One may reason that by restricting the number of possible mutations
in a program; the number of potential errors of the program is also reduced. Testing is essentially
to validate that mutations in the program occur correctly and thus having more mutations would require more testing. By restricting the number of mutations in a program, the program has
fewer reasons for errors to occur and there are fewer state transitions to test.

\subsubsection{Concurrency} Immutable data are thread-safe, as data cannot mutate, there is no danger in having
multiple threads access the same data at the same time and have synchronization issues.

@@

@@

Shadowing is a technique that could represent a changing variable. for instance, an accumulator. the technique is simple and possible in almost every programming language out there.
at it's simplest form shadowing looks like this:
\begin{lstlisting}
    scala> object Main {
        def main(args: Array[String]) = {
            val i = 1;
            {
                val i = 2;
                {
                    val i = 3;
                }
            }
        }  
    }    
\end{lstlisting}
But that's not really useful and even more confusing and very error prone and I'd agree shadowing in of it's self isn't really useful but it's really at the core of any recursive solution since every function come with it's own block and 
\begin{lstlisting} 
    scala> def factorial (n: BigDecimal):BigDecimal = {
        if (n <= 1)  1
        else n * factorial(n-1)
    }
\end{lstlisting}
Notice here n value range over \{n, ... , 1\} but it's not really changing each n deffer from the other and has it's own scope
if you run this code it'll only go so far (around  n = 9613 for this example) until you get a StackOverflowError... not good.

\subsection{Reassignment}
here's a subtle difference, well
hidden behind the overloaded use of the symbol
‘=’, that really sets the two apart. In the imperative
program, ‘=’ refers to a destructive update, assigning
a new value to the left-hand-side variable,
whereas in the functional program ‘=’ means true
equality, and that both the left-hand-side and the right-hand-side can be used interchangeably. This characteristic
of functional programming (known as referential
transparency or purity) has a profound influence
on the way programs are constructed and reasoned
about.

\subsection{Tail Recursion}
Tail recursion provides stack safety to our solutions and prevent Stackoverflows is when you simply return the value of a function call at the end (tail) of your function, in other words your functions has done it's job and handing over the rest of the work to another function
\begin{lstlisting}
  def factorial (n: BigDecimal):BigDecimal = {
    def helper(n: BigDecimal, Acc: BigDecimal): BigDecimal = {
      if (n <= 1) Acc
      else helper(n - 1, n * Acc)
    }
    helper(n, 1)
  }
\end{lstlisting}
Notice that as n takes the values \{n, n-1, n-2, ...\}
the accumulator also changes \{n, n*(n-1), n*(n-1)*(n-2), ...\}
which is very similar to a for loop accumulator pattern, so similar that sometimes compilers compile it to an actual loop

\subsection{Pure Functions}
Pure Functions serves as a (possibly infinite) lookup table mapping from one type to another since variable $x$ will always be the same $f(x)$ will too. this is what's known as referential transparency.
\lstset{style=cpp}
\begin{lstlisting}
    #include <functional>
    #include <iostream>
    int sum(const int v[], const int& n) {
        std::function<int(int, int)> helper = [&v, &n, &helper](int index, int Acc) {
            if (index >= n) return Acc;
            return helper(index + 1, Acc + v[index]);
        };
        return helper(0, 0);
    }
    int main(int, char**) {
        const int a[] = {1, 2, 3, 4, 5, 6};
        // a[1] = 3; not allowed
        int total = sum(a, 6); // 21
        someFunction(a);
        otherFunction(a);
        if (total == sum(a,6))
            std::cout<<"It should be equal same function same argument?!";
        return 0;
    }
\end{lstlisting}
generally speaking $"someFunction"$ and $"otherFunction"$ could've done al sorts of things with $a$ (changing an element value, adding more elements, removing some elements, delete the pointer entirely ...) but if they were pure functions or like in this case using a some sort of a language guarantee (here it's $const$) $a$ will not be modified and in turns $total == sum(a,6)$ and overall our program would be easier to reason about.

\subsection{Laziness}
sometimes referred to as call-by-need, it's the notion that "if data won't change and functions won't neither so would results" meaning that we wouldn't perform any operations unless they're absolutely necessary 
\lstset{style=scala}
\begin{lstlisting}
    def from(n: Int): LazyList[Int] = 
        n #:: from(n+1) 
        // here from(n+1) won't be evaluated
    def sieve (s: LazyList[Int]): LazyList[Int] = 
        s.head #:: sieve(s.tail.filter(_ % s.head != 0)) 
    // nor would sieve(s.tail.filter(_ % s.head != 0))
    val primes = sieve(from(2))

@@

\end{lstlisting}
in this example only 10 elements of the infinite "LazyList" are evaluated while the other elements are immutable they are yet to be evaluated and materialized.
% \subsection{Copy-On-Write (COW)}
% https://en.wikipedia.org/wiki/Copy-on-write
% https://www.researchgate.net/publication/220143982_Effects_of_Copy-on-Write_Memory_Management_on_the_Response_Time_of_UNIX_Fork_Operations

@@

\subsection{Copy-on-Write File Systems}
% COW based file Systems pdf
COW generally follows a simple principle. As long as multiple programs need read only access to a data structure, providing them only pointers[8] which point to the same data structure, instead of copying the whole structure to a new one, is enough. If at least one of the programs needs at some point write access to the data structure, create a private copy for it. The same holds for each one of the programs which will need write access to the data structure. A new private copy will be created for them. Note that the unchanged data can still be shared between both programs. COW file systems also provides many other powerful features, But our main interest for this paper is only one of them which is snapshots.
% Most COW-based file systems provides the ability to take snapshots of your files.

@@

% Leveraging immutability in the implementation snapshots
% https://repositum.tuwien.at/handle/20.500.12708/10774
One of the uses of the snapshot is for enabling of consistent backups. A backup can get inconsistent if during the backup the file system is used. Another use of the snapshot is that the snapshot is kind of easy backup that allows for the retrieving of removed or changed data. Taking of snapshots can be set up in this way that file system could do multiple level undoes. Especially if taking of snapshots does not affect performance of the system.

@@

\newpage
\pagenumbering{Roman}
\setcounter{page}{1}
\section{References}
[1] Berman, J. J. (2018). Immutability and Immortality. In Principles and practice of big data: Preparing, sharing, and analyzing complex information (pp. 80–85). essay, Academic Press. \newline
[2] Trcka, Nikola \& Aalst, Wil \& Sidorova, Natalia. (2009). Data-Flow Anti-patterns: Discovering Data-Flow Errors in Workflows. Proceedings of the International Conference on Advanced Information Systems Engineering (CAiSE). 5565. 425-439. 10.1007/978-3-642-02144-2_34. \newline
[3] Solarwinds: Top 4 Causes Of Storage I/O Bottlenecks \& How To Mitigate Them \newline
[4] Norvig, P. (n.d.). Teach yourself programming in ten years. Retrieved March 5, 2022, from \url{http://norvig.com/21-days.html#answers} \newline
[5] Axelsson, L. (2017). Immutability : An Empirical Study in Scala LUDVIG AXELSSON. \newline 
[6] Smith, Jonathan \& Jr, Gerald. (1988). Effects of Copy-on-Write Memory Management on the Response Time of UNIX Fork Operations.. Computing Systems. 1. 255-278.
[7] Levrinc, R. (2008). LLFS : a copy-on-write file system for Linux [Diploma Thesis]. reposiTUm. \url{https://resolver.obvsg.at/urn:nbn:at:at-ubtuw:1-28018} \newline
[8] Kernighan, B. W., and Ritchie, D. M. The C Programming Lan-
guage, Second Edition. Prentice Hall, 1988. \newline
[9] Narkhede, N., Shapira, G., \& Palino, T. (2017). Meet Kafka. In Kafka: Real-time data and stream processing at scale (pp. 3–5). essay, O'Reilly Media, Incorporated. \newline
[10] Apache Software Foundation. (n.d.). Introduction. Apache Kafka. Retrieved March 10, 2022, from https://kafka.apache.org/intro 
[11]  (n.d.). Retrieved March 11, 2022, from https://elm-lang.org/

>>>>

\title{\Huge{Complex State Management with Immutability}}
\author{Nassouh AlOlabi} 
\affil{Scientific supervisor: Dr.Yasser Rahal}
\affil{General supervisor: Dr.Kadan Joumaa} 
\affil{Linguistic supervisor: Fahmi Alammareen}
\setlength{\affilsep}{1pt}
\date{}

@@

\vspace{1.5cm}
\newpage
\noindent \textbf{Abstract}
\begin{center}
 	\line(1,0){430}
\end{center}
\vspace{-0,3cm}
\noindent Growing demand for fault-tolerant, scalable, distributed systems has made some mainstream software architectures and patterns obsolete or rather harder to come by, Thus came the rise of stateless and functional solutions based on data immutability which has already been the cornerstone of Big Data [1]. We'll take a look at what immutability looks like in a system and what positive outcomes

@@

\vspace{4mm}
% \begin{center}
% 	\line(1,0){430}
% \end{center}

@@

\subsection{Information Flow}
%%https://thedaylightstudio.com/blog/2018/03/14/what-is-state-in-web-application-development
Application State (a.k.a. Program State) represents everything necessary to keep your application running.  When we refer to application state we are normally referring to the state of the program as exists in the contents of its memory.  What does that mean practically?  How am I to understand that?  It helps to think in extremes.  What happens to information and functionality core of your application if a server goes down and restarts?  You lose whatever was residing in memory.

@@

% http://www.padsweb.rwth-aachen.de/wvdaalst/publications/p514.pdf
\subsubsection{Missing Data}
It is the situation where some data element needs to be accessed, i.e. read or destroyed, but either it has never been created or it has been deleted without having been created again.
\subsubsection{Inconsistent Data} 
Data is inconsistent if a task is using this data while some other task (or another instance of the same task) is writing to this data or is destroying it in parallel.
%-----------------------------
\subsection{Programming Paradigms}
% https://www.geeksforgeeks.org/introduction-of-programming-paradigms/
Programming paradigm is an approach to solve problems using some programming languages. It is also a method to solve a problem using available tools and techniques.
%----------------------------------------
Most paradigms have their own opinion on data manipulation for example: Procedural \& Object Oriented paradigms allow more access to data modifying (destructive updates), deleting. Others like Logic \& Functional does not allow these modifications and deletions.

\subsection{I/O, Network and Disk}
% https://www.solarwinds.com/-/media/solarwinds/swdcv2/licensed-products/storage-resource-monitor/resources/whitepapers/top_4_causes_of_storage_io_bottlenecks.ashx
Applications that are I/O heavy often cause bottlenecks. I/O intensive applications are more sensitive to a storage latency issue. When you have a large user base trying to access these applications, slowdowns tend to take place. Increased response time in storage I/O causes bottlenecks. When there is a queue in the storage I/O, you would generally see an increase in latency.
% TODO: often times
Network and Disk latency are often times the biggest source of headache if poorly managed, which is often times the case since they're usually treated as if they were function calls which they're clearly not [3].
% http://norvig.com/21-days.html#answers
\begin{center}
    \begin{table}[H]
        \begin{center}
        \begin{tabularx} {0.8\textwidth}{ 
            | >{\raggedright\arraybackslash}X 
        | >{\raggedleft\arraybackslash}X | }
        \hline
        execute typical instruction &	1/1,000,000,000 sec = 1 nanosec\\
\hline
    fetch from L1 cache memory &	0.5 nanosec\\
\hline
branch misprediction &	5 nanosec\\
\hline
    fetch from L2 cache memory &	7 nanosec\\
\hline
    Mutex lock/unlock &	25 nanosec\\
\hline
    fetch from main memory &	100 nanosec\\
\hline
    send 2K bytes over 1Gbps network &	20,000 nanosec\\
    \hline
    read 1MB sequentially from memory &	250,000 nanosec\\
\hline
    fetch from new disk location (seek) &	8,000,000 nanosec\\
\hline
    read 1MB sequentially from disk &	20,000,000 nanosec\\
\hline
    send packet US to Europe and back &	150 milliseconds = 150,000,000 nanosec\\
    \hline      
    % \float
\end{tabularx}

\end{center}
\caption{ \label{tab:1} Latency numbers every programmer should know[4]}
\end{table}
\end{center}
As you can see disk fetches can take 100x more than main memory fetches. When dealing with distributed systems where network call are integral parts of the system, treating network calls which are 100000x more than main memory calls ends up with catastrophically unpredictable behavior

@@

% src
% Immutability an Empirical Study in Scala.pdf
When something is immutable, we say that it cannot be changed or that it is unchangeable. An immutable object in an object-oriented programming language is an object that has a state that cannot be mutated once instantiated (created). Moreover, an immutable class is a class whose instances cannot be mutated, meaning that there are no methods in the class that can mutate an instance of it. Classes that are not immutable can have instances that are immutable.

\subsection{Forms of Immutability}
There are different varieties of immutability used today. They includes:
\begin{itemize}
    \item Object immutability
    \begin{description}
        \item[] An immutable object is an object that cannot be modified (mutated), i.e., its state cannot be mutated
    \end{description}
    \item Class immutability 
    \begin{description}
        \item[] When every instantiated object of a class is immutable, then we say that the class itself is immutable.
    \end{description}
    \item Deep and shallow immutability (transitivity)\begin{description}
        \item[] Immutability can be deep or shallow, i.e., transitive or non-transitive:
        \item[]Deep (transitive): immutability means that all objects referred to by an immutable object must also be immutable.
        \item[]Shallow (non-transitive) immutability has a more relaxed constraint, and the immutable object may refer to objects that are mutable, but the fields of the immutable object itself cannot be mutated.
    \end{description}    
    \item Reference immutability (read-only references)
     \begin{itemize}
    \item[] Languages with the support of reference immutability have the notion of read-only references. A readonly reference cannot mutate the object it refers to. When all references to an object are readonly, then nothing can mutate the object, and the object is immutable. There is no guarantee that a referred to object is immutable because the object may still be mutated by some other references that are not read-only, unless there is some analysis that ensures that there only exist readonly references to that object. Reference immutability thus often ensures shallow (non-transitive) immutability and “reference immutability” does not mean that a reference itself is immutable [5].
\end{itemize}

\item Non-assignability
     \begin{itemize}
    \item[] Non-assignability is a form of immutability and property of a variable that makes sure that the variable cannot be reassigned. Since fields of objects are variables and if no variable can be reassigned after its initial assignment, the object is effectively immutable if what the fields are referring to is immutable. This makes non-assignability give shallow (non-transitive) immutability too. Assignment of an object’s field mutates the object, but it does not mutate what was previously on the field or what was assigned to the field.
\end{itemize}

\item Concrete and abstract immutability
     \begin{itemize}
    \item[] Concrete immutability does not allow any change to the object’s state. Abstract immutability, however, allows an immutable object to change its internal state but not the object’s “abstract” value. The object is still immutable from the perspective of the object’s observer, who can only see the abstract value, but the object may mutate internally. This can, for example, be useful to speed up certain operations, lazy initialization and buffering.
\end{itemize}

@@

\subsection{Properties of Immutability}
% write some problems
% The concept of immutability is important and
Utilizing immutable data is considered to ease software development and reasoning about programs in
numerous ways, for example:

\subsubsection{Predictability} It is hard to understand and reason about programs that have shared mutable
states with unclear interactions. Tracking mutations and maintaining the correct state of a program
can be difficult. Using immutable data naturally, avoids state changes and forces the programmer
to let data flow and be utilized in a different way throughout the program, making the state of the
program more predictable. For example, calling the same function twice would yield the same result,
and the outcome is predictable. Without the immutability guarantee, the second call could
yield another result because of an underlying state mutation making it less predictable.

\subsubsection{Testability} Because immutable data can only be changed once during construction, they are inherently
simpler and easier to unit test. One may reason that by restricting the number of possible mutations
in a program; the number of potential errors of the program is also reduced. Testing is essentially
to validate that mutations in the program occur correctly and thus having more mutations would require more testing. By restricting the number of mutations in a program, the program has
fewer causes of errors to occur and there will be fewer state transitions to test.

\subsubsection{Concurrency} Immutable data are thread-safe. As data cannot mutate, there is no danger in having
multiple threads access the same data at the same time and have synchronization issues.

@@

\subsection{Reassignment}
There's a subtle difference, well
hidden behind the overloaded use of the symbol
‘=’, that really sets the two apart. In the imperative
program, ‘=’ refers to a destructive update; assigning
a new value to the left-hand-side variable,
whereas in the functional program ‘=’ means true
equality, and that both the left-hand-side and the right-hand-side can be used interchangeably. This characteristic
of functional programming (known as referential
transparency or purity) has a profound influence
on the way programs are constructed and reasoned
about.

@@

Shadowing is a technique that could be used to represent a changing variable, like an accumulator. This technique is simple and possible in almost every programming language.
At its simplest form, shadowing looks like this:
\begin{lstlisting}
    scala> object Main {
        def main(args: Array[String]) = {
            val i = 1;
            {
                val i = 2;
                {
                    val i = 3;
                }
            }
        }  
    }    
\end{lstlisting}
But that's not really useful and even more confusing and very error prone. You are right. shadowing in itself isn't really useful. but it is really at the core of any recursive solution since every function come with its own block and 
\begin{lstlisting} 
    scala> def factorial (n: BigDecimal):BigDecimal = {
        if (n <= 1)  1
        else n * factorial(n-1)
    }
\end{lstlisting}
Notice here n value ranges over \{n, ... , 1\} but it isn't really changing... each n deffer from the other and has it's own scope
if you run this code it'll only go so far (around  n = 9613 for this example) until you get a StackOverflowError... not good.

\subsection{Tail Recursion}
Tail recursion provides stack safety to our solutions and prevents Stackoverflows. It is when you simply return the value of a function call at the end (tail) of your function. in other words, your function has done its job and hands over the rest of the work to another function.
\begin{lstlisting}
  def factorial (n: BigDecimal):BigDecimal = {
    def helper(n: BigDecimal, Acc: BigDecimal): BigDecimal = {
      if (n <= 1) Acc
      else helper(n - 1, n * Acc)
    }
    helper(n, 1)
  }
\end{lstlisting}
Notice that as n takes the values \{n, n-1, n-2, ...\}
the accumulator also changes \{n, n*(n-1), n*(n-1)*(n-2), ...\}
which is very similar to a for loop accumulator pattern, so similar that sometimes compilers compile tail recursive functions to an actual loop

\subsection{Pure Functions}
Pure Functions serve as a (possibly infinite) lookup table mapping from one type to another since variable $x$ will always be the same, $f(x)$ will be the same too. this is what's known as referential transparency.
\lstset{style=cpp}
\begin{lstlisting}
    #include <functional>
    #include <iostream>
    int sum(const int v[], const int& n) {
        std::function<int(int, int)> helper = [&v, &n, &helper](int index, int Acc) {
            if (index >= n) return Acc;
            return helper(index + 1, Acc + v[index]);
        };
        return helper(0, 0);
    }
    int main(int, char**) {
        const int a[] = {1, 2, 3, 4, 5, 6};
        // a[1] = 3; not allowed
        int total = sum(a, 6); // 21
        someFunction(a);
        otherFunction(a);
        if (total == sum(a,6))
            std::cout<<"It should be equal same function same argument?!";
        return 0;
    }
\end{lstlisting}
generally speaking $"someFunction"$ and $"otherFunction"$ could've done all sorts of things with $a$ (changing an element value, adding more elements, removing some elements, delete the pointer entirely ...). If they were pure functions or like in this case using some sort of a language guarantee (here it's $const$) $a$ will not be modified and in turn $total == sum(a,6)$ and overall, our program would be easier to reason about.

\subsection{Laziness}
sometimes referred to as call-by-need, it's the notion that "if data won't change neither will function outputs " meaning that we wouldn't perform any operation unless it is absolutely necessary 
\lstset{style=scala}
\begin{lstlisting}
    def from(n: Int): LazyList[Int] = 
        n #:: from(n+1) 
        // here from(n+1) won't be evaluated
    def sieve (s: LazyList[Int]): LazyList[Int] = 
        s.head #:: sieve(s.tail.filter(_ % s.head != 0)) 
    // nor would sieve(s.tail.filter(_ % s.head != 0))
    val primes = sieve(from(2))

@@

\end{lstlisting}
In this example only 10 elements of the infinite "LazyList" are evaluated. The other elements are immutable they are yet to be evaluated and materialized.
% \subsection{Copy-On-Write (COW)}
% https://en.wikipedia.org/wiki/Copy-on-write
% https://www.researchgate.net/publication/220143982_Effects_of_Copy-on-Write_Memory_Management_on_the_Response_Time_of_UNIX_Fork_Operations

@@

\subsection{Copy-on-Write File Systems}
% COW based file Systems pdf
COW generally follows a simple principle. As long as multiple programs need read only access to a data structure, providing them only pointers[7] which point to the same data structure, instead of copying the whole structure to a new one, is enough. If at least one of the programs needs at some point write access to the data structure, create a private copy for it. The same holds for each one of the programs which will need write access to the data structure. A new private copy will be created for them. Note that the unchanged data can still be shared between both programs. COW file systems also provides many other powerful features, But our main interest for this paper is only one of them which is snapshots.
% Most COW-based file systems provides the ability to take snapshots of your files.

@@

% Leveraging immutability in the implementation snapshots
% https://repositum.tuwien.at/handle/20.500.12708/10774
One of the uses of the snapshot is for enabling of consistent backups. A backup can get inconsistent if during the backup the file system is used. Another use of the snapshot is that the snapshot is kind of easy backup that allows for the retrieving of removed or changed data. Taking of snapshots can be set up in this way that file system could do multiple level undoes. Especially if taking of snapshots does not affect performance of the system[8].

@@

\newpage
\pagenumbering{Roman}
\setcounter{page}{1}
\section{References}
[1] Berman, J. J. (2018). Immutability and Immortality. In Principles and practice of big data: Preparing, sharing, and analyzing complex information (pp. 80–85). essay, Academic Press. \newline
[2] Trcka, Nikola \& Aalst, Wil \& Sidorova, Natalia. (2009). Data-Flow Anti-patterns: Discovering Data-Flow Errors in Workflows. Proceedings of the International Conference on Advanced Information Systems Engineering (CAiSE). 5565. 425-439. 10.1007/978-3-642-02144-2_34. \newline
[3] Solarwinds: Top 4 Causes Of Storage I/O Bottlenecks \& How To Mitigate Them \newline
[4] Norvig, P. (n.d.). Teach yourself programming in ten years. Retrieved March 5, 2022, from \url{http://norvig.com/21-days.html#answers} \newline
[5] Axelsson, L. (2017). Immutability : An Empirical Study in Scala LUDVIG AXELSSON. \newline 
[6] Smith, Jonathan \& Jr, Gerald. (1988). Effects of Copy-on-Write Memory Management on the Response Time of UNIX Fork Operations.. Computing Systems. 1. 255-278.
[7] Kernighan, B. W., and Ritchie, D. M. The C Programming Lan-
guage, Second Edition. Prentice Hall, 1988. \newline
[8] Levrinc, R. (2008). LLFS : a copy-on-write file system for Linux [Diploma Thesis]. reposiTUm. \url{https://resolver.obvsg.at/urn:nbn:at:at-ubtuw:1-28018} \newline
[9] Narkhede, N., Shapira, G., \& Palino, T. (2017). Meet Kafka. In Kafka: Real-time data and stream processing at scale (pp. 3–5). essay, O'Reilly Media, Incorporated. \newline
[10] Apache Software Foundation. (n.d.). Introduction. Apache Kafka. Retrieved March 10, 2022, from https://kafka.apache.org/intro \newline
[11]  (n.d.). Retrieved March 11, 2022, from https://elm-lang.org/
